---
title: "R-Ladies Newcastle Christmas Social"
output:
  xaringan::moon_reader:
    css: ["default", "rladies", "rladies-fonts"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
# A few fun facts

We thought we'd take this chance to look back at R-Ladies Newcastle first year. 

*  __7 meetups in 7 months!__
* __211 members on Meetup!__

```{r setup, include = FALSE}
library(xts)
library(magrittr)
library(ggplot2)
library(dplyr)
library(reshape2)
library(tidyverse)
library(lubridate)
library(zoo)

timeseries_counts <- read.csv("timeseries_counts.csv", header = TRUE)

timeseries_counts$Date <- as.Date(timeseries_counts$Date, format = "%d/%m/%Y")
```


---
# Organisers

We started this year with 6 organisers. 

We currently have 4 organisers, but would love an additional two organisers. 

We're a friendly bunch and you can make the role as large as you want. It could be as small as just helping us welcome people to the meetup.

---
# Meetup Attendance
```{r attendance, echo = FALSE}
attendee_plot <- fortify(timeseries_counts) %>%
   ggplot(aes(x = Date, y = Attendee.count, group = 1))+   
  geom_line() +
  xlab("Date") +
  ylab("") +
  theme_classic()+
  theme(legend.position = "none")

print(attendee_plot)
```
---
# How many pizzas have been eaten?

```{r pizza, echo = FALSE}

pizza_plot <- fortify(timeseries_counts) %>%
   ggplot(aes(x = Date, y = Pizza.count, group = 1))+   
  geom_line() +
  xlab("Date") +
  ylab("") +
  theme_classic()+
  theme(legend.position = "none")

print(pizza_plot)

```
---
#FYI: SatRdays

We encourage fellow R-Ladies to submit talks!

If you're nervous about submitting an abstract- Aoife would be happy to proof read or if you want to practice your talk before hand- is also happy to offer her services.

https://newcastle2020.satrdays.org 

---
# Feedback and chat

Please share your thoughts!

Have a mince pie (they're vegan!)

Chat to someone you've not talked too before
