---
title: "Data Visualisation in R"
author: "Aoife, Jess, Rike"
output: ioslides_presentation
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, error = F)
library(gapminder)
library(tidyverse)
library(gganimate)
```


## Learning outcomes

- Make different types of plots such as scatter plots, bar plots, histograms, box plots, area plots and density plot
- Easily make subplots of your data with faceting 
- Arrange plots in a grid
- Customise your plots

## How do GDP and life expectancy change over time?

## 

```{r plot, echo = F, fig.height = 3.3, fig.width = 4.2}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp,  size = pop, colour = continent)) +
  geom_point(alpha = .7) +
  facet_wrap(~ continent) +
  scale_colour_brewer(palette = "Dark2") +
  scale_size(name = "Population") +
  guides(colour = "none") +
  scale_x_log10() +
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'Life expectancy') +
  transition_time(year)
```


## Layers

```ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) + 
  <GEOM_FUNCTION>()```
  
## Data, x, y

``` {r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
```

## Type of plot

``` {r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +  
  geom_point()
```

## Colour and size

``` {r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp,  
                             colour = continent, size = pop)) +  
  geom_point()
```

## Adjusting the colour, size, etc: scales

``` {r fig.height = 4}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp,  
                             colour = continent, size = pop)) +  
  geom_point() +  
  scale_colour_brewer(palette = "Dark2")
```



## Making subplots: facets

``` {r fig.height = 3.7}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp,  
                             colour = continent, size = pop)) +  
  geom_point() +  
  scale_colour_brewer(palette = "Dark2") +  
  facet_wrap(~ continent)
```

## Adjusting the axis titles

``` {r fig.height = 3.5}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp,  
                             colour = continent, size = pop)) +  
  geom_point() +  
  scale_colour_brewer(palette = "Dark2") +  
  facet_wrap(~ continent) +  
  labs(x = "GDP per capita", y = "Life expectancy")
```

## Adjusting the overall plot look: theme

``` {r fig.height = 3.2}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp,  
                             colour = continent, size = pop)) +  
  geom_point() +  
  scale_colour_brewer(palette = "Dark2") +  
  facet_wrap(~ continent) +  
  labs(x = "GDP per capita", y = "Life expectancy") +
  theme_classic()
```

## Useful links/resources


There is lots of useful info online on making plots with ggplot2, but this book is a great starting point:

Wickham, H. and Grolemund, G. (2016) R for data science. Free at http://r4ds.had.co.nz/ It was written by the guy who wrote the package ggplot2. 

Another useful book is this one: http://www.cookbook-r.com/Graphs/